Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-03 talk/dls2014/paper/paper.tex
clean up cites and a few tweaks

TM systems can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} -- sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

@@

On encountering conflicts, we employ contention management to solve
the problem as well as we can. The general rules are:
\begin{itemize}[noitemsep]
\item prefer transactions that started earlier to younger transactions
\item to support \emph{inevitable} transactions, we always prefer them
  to others since they cannot abort (similar to \cite{blundell06})
\end{itemize}
We can either simply abort a transaction to let the other one succeed,
or we can also wait until the other transaction committed. The latter
is an interesting option if we are trying to commit a write and
another transaction already read the object. We can then signal the
other transaction to commit as soon as possible and wait. After
waiting, there is now no conflict between our write and the already
committed read anymore.

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in Section~\ref{sec:performance-bench}. We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT is a tracing JIT\cite{cfbolz09}
\remi{explain?}  running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

@@

As expected, all interpreters with a GIL do not scale with the number
of threads. They even become slower because of the overhead of
thread-switching and GIL handling (see \cite{beazley10} for a detailed
analysis). We also see Jython scale when we expect it to (\emph{mandelbrot,
raytrace, richards}), and behave similar to the GIL interpreters in the
other cases. The reason again being the coarse grained locking.

@@

Other attempts using HTM~\cite{nicholas06,odaira14,fuad10} show
promising results, especially the very recent \cite{odaira14} that
examines the current generation of Intel CPUs. However, they did a
simplification by essentially disabling the GC during the
benchmarks. This may be valid for the hand-tuned allocations in their
interpreter. For us the GC is essential, and because it is a moving
GC, it unfortunately touches a lot more memory and thereby reaches the
HTM capacity much faster.  This, and the fact that HTM's capacity
limits do not allow the implementation of arbitrary atomic blocks, let
us to believe that STM is the right choice at the moment.

There are multiple attempts at removing the limits of HTM by means of
\emph{virtualising} HTM~\cite{rajwar05,chung06}. Indeed this may also
be the future direction of our research. So far, some of them depend
on additional hardware features. Or in the case of \cite{chung06},
they have the assumption that most transactions fit the HTM capacity
and that virtualisation is only a fallback. Our use case, however,
features arbitrarily long transactions, which is why this assumption
first has to be verified in our case.

>>>>

TM systems can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In~\cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} -- sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

@@

On encountering conflicts, we employ contention management to solve
the problem as well as we can. The general rules are:
\begin{itemize}[noitemsep]
\item prefer transactions that started earlier to younger transactions
\item to support \emph{inevitable} transactions, we always prefer them
  to others since they cannot abort (similar to~\cite{blundell06})
\end{itemize}
We can either simply abort a transaction to let the other one succeed,
or we can also wait until the other transaction committed. The latter
is an interesting option if we are trying to commit a write and
another transaction already read the object. We can then signal the
other transaction to commit as soon as possible and wait. After
waiting, there is now no conflict between our write and the already
committed read anymore.

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in Section~\ref{sec:performance-bench}. We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT is a tracing JIT~\cite{cfbolz09}
\remi{explain?}  running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

@@

As expected, all interpreters with a GIL do not scale with the number
of threads. They even become slower because of the overhead of
thread-switching and GIL handling (see~\cite{beazley10} for a detailed
analysis). We also see Jython scale when we expect it to (\emph{mandelbrot,
raytrace, richards}), and behave similar to the GIL interpreters in the
other cases. The reason again being the coarse grained locking.

@@

Other attempts using HTM~\cite{nicholas06,odaira14,fuad10} show
promising results, especially the very recent~\cite{odaira14} that
examines the current generation of Intel CPUs. However, they did a
simplification by essentially disabling the GC during the
benchmarks. This may be valid for the hand-tuned allocations in their
interpreter. For us the GC is essential, and because it is a moving
GC, it unfortunately touches a lot more memory and thereby reaches the
HTM capacity much faster.  This, and the fact that HTM's capacity
limits do not allow the implementation of arbitrary atomic blocks, let
us to believe that STM is the right choice at the moment.

There are multiple attempts at removing the limits of HTM by means of
\emph{virtualising} HTM~\cite{rajwar05,chung06}. Indeed this may also
be the future direction of our research. So far, some of them depend
on additional hardware features. Or in the case of~\cite{chung06},
they have the assumption that most transactions fit the HTM capacity
and that virtualisation is only a fallback. Our use case, however,
features arbitrarily long transactions, which is why this assumption
first has to be verified in our case.
